A tutorial of Bayesian beta regressions with brms in R

### 2025-02-05 manuscript/manuscript.qmd
use tables in manuscript

library(tidyverse)
library(brms)
library(collapse)
```

@@

```{r}
#| label: voi-prop-fixef
#| tbl-cap: "Regression coefficients of a beta regression of voicing proportion (`voi_prop_bm`)."
#| eval: false

@@

Once we have the posterior difference, we can obtain CrIs of the difference using `quantile2()` from the posterior package. Beware that the values of the difference are in log-odds! We can convert these into odd-ratios with `exp()`. odd-ratios indicate the ratio of the difference between A and B, so that 1 means no difference, values greater than 1 indicate an increase in A relative to B and values lower than 1 indicate a decrease in A relative to B. odd-ratios are useful when looking at differences that are in log-odds because while the relative magnitude of the difference in proportion between two groups is the same independent of the baseline proportion, the *absolute* magnitude of the difference depends on the baseline value. For example, an odd-ratio difference of 1.25 would correspond to a proportion increase of 13 percentage points if the baseline proportion is 0.62 but it would correspond to a proportion increase of 17 percentage points if the baseline proportion is 0.73. Of course, in real research contexts it is still usefull to think about absolute magnitudes and their relevance from a conceptual and methodological perspective. In this tutorial we just focus on odd-ratios for simplicity.

```{r}
#| label: voi-prop-bm-diff-cri

@@

voi_prop_bm_diff |> 
  mutate(c2p_tk_diff_ratio = exp(c2p_tk_diff)) |> 
  reframe(
    # 90% CrI
    q90 = quantile2(c2p_tk_diff_ratio, probs = c(0.05, 0.95)),
    # 80% CrI
    q80 = quantile2(c2p_tk_diff_ratio, probs = c(0.1, 0.9)),
    # 60% CrI
    q60 = quantile2(c2p_tk_diff_ratio, probs = c(0.2, 0.8)),
  ) |> 
  # round to 2 digits
  mutate(across(everything(), ~round(.x, 2)))
```

So, based on the model and data, there is a 90% probability that the voicing proportion in /p/ is 1.07-1.38 times longer (or 7-38% increase) than in /t, k/. At 80% confidence, the change ratio is 1.10-1.34 (or 10-34% increase) while at 60% confidence is 1.14-1.30 (14-30% increase). In other words we can be quite confident that the voicing proportion in /p/ is longer than in /t, k/ and that the increase is less than 35%.

brms comes with a convenient function, `conditional_effects()`, to plot posterior means and CrIs based on predictors in the model. In the following example, we plot the predicted proportion of voicing by consonant and vowel.

```{r}
#| label: voi-prop-bm-cond
conditional_effects(voi_prop_bm, "c2:vowel")
```

@@

```{r}
#| label: marginal

@@

@@

First, let's read and plot the data.

nasal <- read_csv("data/carignan2021/nasal.csv")
nasal
```

@@

@@

Let's inspect the output of `fixef()`.

```{r}
#| label: nas-prop-bm-fixef
round(
  fixef(nas_prop_bm, prob = c(0.1, 0.9)),
  digits = 2
)
```

Negative log-odds indicate a proportion that is smaller than 50%, while positive log-odds a proportion that is greater than 50%. Generally, the expected log-odd predictions are negative, indicating an overall tendency for the nasalisation to take less than 50% of the duration of the vowel. Moreover, the predictions are higher for voiceless NC sequences than for voiced NC sequences, indicating a greater proportion of nasalisation in the former. However there is vowel-specific variation, and there doesn't seem to be much of a difference in nasalisation proportion in /e/ and /u/. Plotting the expected predictions with `conditional_effects()` should make this clearer.

```{r}
#| label: nas-prop

@@

To answer question 1, we can calculate the average difference in nasalisation proportion by first calculating the average nasalisation across all vowels for voiced and voiceless sequences and then take the difference of those, similarly to what we have done in the Case Study 1 above.

```{r}
#| label: nas-prop-bm-draws

@@

nas_prop_bm_draws_long <- nas_prop_bm_draws |> 
  # pivot vowel:c2 columns
  pivot_longer(`b_voicingvoiced:vowela`:`b_voicingvoiceless:vowelu`, names_to = "coeff") |>
  # separate "coeff" labels into type ("b"), vowel and c2
  separate(coeff, into = c("type", "voicing", "vowel"))

nas_prop_bm_draws_long
```

Now let's calculate the mean nasalisation proportion within each draw by voicing, and plot the resulting posterior distributions. Note that, as discussed for Case study 1, when working wit log-odds it is important to first do all necessary calculations in log-odds, here calculate the mean log-odds across vowels, and *then* transform the calculated estimands to proportions/probabilities.

@@

nas_prop_bm_draws_long_voicing <- nas_prop_bm_draws_long |> 
  # grouping by .draw and voicing ensures that averaging applies only within draw and voicing
  group_by(.draw, voicing) |> 
  summarise(
    # calculate the mean value within draw/voicing in log-odds
    value_mean = mean(value),
    # we can now transform log-odds to proportion with plogis()
    value_mean_prop = plogis(value_mean),
    .groups = "drop"
  )

nas_prop_bm_draws_long_voicing
```

```{r}
#| label: nas-prop-bm-draws-long-voicing-plot

@@

The plot suggests an overall greater nasalisation proportion in voiceless NC sequences. Let's quantify how greater as we did in Case Study 1. We will use odd-ratios in this context as well, i.e. we will convert log-odds to odd-ratios using the `exp()` (exponential) function (and as before we first calculate the difference and then exponentiate the resulting values, after which we can take summary measures, like means and quantile-based measures such as CrIs).

@@

nas_prop_bm_diff <- nas_prop_bm_draws_long_voicing |> 
  # pivot data to create one column per voicing with the mean drawn values,
  # with one draw per raw. we need to drop the value_mean_prop col
  select(-value_mean_prop) |> 
  pivot_wider(names_from = voicing, values_from = value_mean) |> 
  mutate(
    # calculate the difference of voiceless and voiced in log-odds
    voicing_diff = voicingvoiceless - voicingvoiced,
    # now transform with exp() to get the ratio difference
    voicing_diff_ratio = exp(voicing_diff)
  )
nas_prop_bm_diff
```

```{r}
#| label: nas-prop-bm-diff-cris

nas_prop_bm_diff |> 
  reframe(
    # 90% CrI
    q90 = quantile2(voicing_diff_ratio, probs = c(0.05, 0.95)),
    # 80% CrI
    q80 = quantile2(voicing_diff_ratio, probs = c(0.1, 0.9)),
    # 60% CrI
    q60 = quantile2(voicing_diff_ratio, probs = c(0.2, 0.8)),
  ) |> 
  mutate(across(everything(), ~round(.x, 2)))
```

The CrIs of the ratio difference in nasalisation proportion in voiceless vs voiced NC sequences suggest a robust increase of nasalisation in the voiceless NC sequences, with a 90% probability that the increase is between 23% and 69% of the proportion in voiced NC sequences.

@@

nas_prop_r <- nas_prop_bm |> 
  # extract draws from model, only `r_speaker` varying terms
  spread_draws(r_speaker[speaker,voicingvowel]) |> 
  # separate the column voicingvowel to two columns
  separate(voicingvowel, c("voicing", "vowel")) |> 
  # join the draws with the `b_` terms
  left_join(y = nas_prop_bm_draws_long) |> 
  # get the expected log-odd value of each speaker, in each draw
  # this is the sum of the `value` from the b_ terms and the value from the
  # r_speaker term.
  mutate(r_speaker_value = value + r_speaker) |> 
  # group the data for summarise
  group_by(.draw, speaker, voicing) |> 
  # get mean expected log-odds by draw, speaker and voicing (averaging across vowel)
  summarise(r_speaker_value_mean = mean(r_speaker_value)) |> 
  # make the data wider: two columns, one for voiced and one for voiceless
  pivot_wider(names_from = voicing, values_from = r_speaker_value_mean) |> 
  # finally, calculate the difference in expected log-odds of voiceless and voiced
  mutate(voicing_diff = voicingvoiceless - voicingvoiced)

nas_prop_r
```

@@

```{r}
#| label: fig-nas-prop-r
#| warning: false
#| fig-asp: 1.5

>>>>

library(tidyverse)
library(brms)
library(collapse)
library(marginaleffects)
```

@@

```{r}
#| label: voi-prop-fixef
#| eval: false

@@

Once we have the posterior difference, we can obtain CrIs of the difference using `quantile2()` from the posterior package. Beware that the values of the difference are in log-odds! We can convert these into odd-ratios with `exp()`. The output of the code below is shown in @tbl-voi-bm-quant. Odd-ratios indicate the ratio of the difference between A and B, so that 1 means no difference, values greater than 1 indicate an increase in A relative to B and values lower than 1 indicate a decrease in A relative to B. odd-ratios are useful when looking at differences that are in log-odds because while the relative magnitude of the difference in proportion between two groups is the same independent of the baseline proportion, the *absolute* magnitude of the difference depends on the baseline value. For example, an odd-ratio difference of 1.25 would correspond to a proportion increase of 13 percentage points if the baseline proportion is 0.62 but it would correspond to a proportion increase of 17 percentage points if the baseline proportion is 0.73. Of course, in real research contexts it is still usefull to think about absolute magnitudes and their relevance from a conceptual and methodological perspective. In this tutorial we just focus on odd-ratios for simplicity.

```{r}
#| label: voi-prop-bm-diff-cri
#| message: false

@@

voi_bm_quant <- voi_prop_bm_diff |> 
  mutate(c2p_tk_diff_ratio = exp(c2p_tk_diff)) |> 
  reframe(
    # 90% CrI
    q90 = quantile2(c2p_tk_diff_ratio, probs = c(0.05, 0.95)),
    # 80% CrI
    q80 = quantile2(c2p_tk_diff_ratio, probs = c(0.1, 0.9)),
    # 60% CrI
    q60 = quantile2(c2p_tk_diff_ratio, probs = c(0.2, 0.8)),
  ) |> 
  # round to 2 digits
  mutate(across(everything(), ~round(.x, 2)))
```

```{r}
#| label: tbl-voi-bm-quant
#| tbl-cap: "Upper and lower limits of 90, 80 and 60% Credible Intervals of the difference ratio of voicing proportion in /t, k/ vs /p/."
#| echo: false

voi_bm_quant |> knitr::kable()
```

Based on the model and data, there is a 90% probability that the voicing proportion in /p/ is 1.07-1.38 times longer (or 7-38% increase) than in /t, k/. At 80% confidence, the change ratio is 1.10-1.34 (or 10-34% increase) while at 60% confidence is 1.14-1.30 (14-30% increase). In other words we can be quite confident that the voicing proportion in /p/ is longer than in /t, k/ and that the increase is less than 35%. The brms package comes with a convenient function, `conditional_effects()`, to plot posterior means and CrIs based on predictors in the model. In @fig-voi-prop-bm-cond, we plot the predicted proportion of voicing by consonant and vowel.

```{r}
#| label: fig-voi-prop-bm-cond
#| fig-cap: "Expected voicing proportion by consonant and vowel with 95% Credible Intervals."
conditional_effects(voi_prop_bm, "c2:vowel")
```

@@

```{r}
#| label: marginal-code
#| eval: false

@@

```{r}
#| label: tbl-marginal-1
#| tbl-cap: "Difference in expected voicing proportion by consonant, averaged across vowels, with 80% CrIs."
#| echo: false

avg_comparisons(voi_prop_bm, variables = list(c2 = "pairwise"), conf_level = 0.8, type = "link") |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  select(-(predicted_lo:tmp_idx)) |> 
  knitr::kable()
```

```{r}
#| label: tbl-marginal-2
#| tbl-cap: "Expected voicing proportion by vowel, averaged across vowels, with 80% CrIs."
#| echo: false

avg_predictions(voi_prop_bm, variables = "vowel", conf_level = 0.8) |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  knitr::kable()
```

@@

@tbl-nasal shows the first ten rows of the data. The data contains the following columns:

@@

nasal <- read_csv("data/carignan2021/nasal.csv")
```

```{r}
#| label: tbl-nasal
#| tbl-cap: "First ten rows of the nasal proportion data."
#| echo: false

nasal |> head(n = 10) |> knitr::kable()
```

@@

Let's inspect the output of `fixef()`, reported in @tbl-nas-prop-fixef.

```{r}
#| label: nas-prop-bm-fixef
#| eval: false

fixef(nas_prop_bm, prob = c(0.1, 0.9))
```

```{r}
#| label: tbl-nas-prop-fixef
#| tbl-cap: "Regression coefficients of a beta regression of nasalisation proportion (`nas_prop_bm`)."
#| echo: false

round(
  fixef(nas_prop_bm, prob = c(0.1, 0.9)),
  digits = 2
) |> knitr::kable()
```

Negative log-odds indicate a proportion that is smaller than 50%, while positive log-odds a proportion that is greater than 50%. Generally, the expected log-odd predictions are negative, indicating an overall tendency for the nasalisation to take less than 50% of the duration of the vowel. Moreover, the predictions are higher for voiceless NC sequences than for voiced NC sequences, indicating a greater proportion of nasalisation in the former. However there is vowel-specific variation, and there doesn't seem to be much of a difference in nasalisation proportion in /e/ and /u/. @fig-nas-prop shows the expected predictions with `conditional_effects()`, which should make the interpretation clearer.

```{r}
#| label: fig-nas-prop
#| fig-cap: "Expected nasalisation proportion by vowel and voicing, with 95% CrIs."

@@

To answer question 1, we can calculate the average difference in nasalisation proportion by first calculating the average nasalisation across all vowels for voiced and voiceless sequences (see @tbl-nas-draws for the output of this step) and then take the difference of those, similarly to what we have done in the Case Study 1 above.

```{r}
#| label: nas-prop-bm-draws
#| warning: false

@@

nas_prop_bm_draws_long <- nas_prop_bm_draws |> 
  # pivot vowel:c2 columns
  pivot_longer(`b_voicingvoiced:vowela`:`b_voicingvoiceless:vowelu`, names_to = "coeff") |>
  # separate "coeff" labels into type ("b"), vowel and c2
  separate(coeff, into = c("type", "voicing", "vowel"))
```

```{r}
#| label: tbl-nas-draws
#| tbl-cap: "First ten rows of posterior draws from `nas_prop_bm` in long format."
#| echo: false

nas_prop_bm_draws_long |> head(n = 10) |> knitr::kable()
```

Now let's calculate the mean nasalisation proportion within each draw by voicing, and plot the resulting posterior distributions. Note that, as discussed for Case study 1, when working wit log-odds it is important to first do all necessary calculations in log-odds, here calculate the mean log-odds across vowels, and *then* transform the calculated estimands to proportions/probabilities. The output of the following code is show in @tbl-nas-long-voi and the density plot of the calculated draws is in @fig-nas-prop-bm-draws-long-voicing-plot.

@@

nas_prop_bm_draws_long_voicing <- nas_prop_bm_draws_long |> 
  # grouping by .draw and voicing ensures that averaging applies only within draw and voicing
  group_by(.draw, voicing) |> 
  summarise(
    # calculate the mean value within draw/voicing in log-odds
    value_mean = mean(value),
    # we can now transform log-odds to proportion with plogis()
    value_mean_prop = plogis(value_mean),
    .groups = "drop"
  )
```

```{r}
#| label: tbl-nas-long-voi
#| tbl-cap: "Mean nasalisation proportion by draw and voicing, averaged across vowels (first 10 rows)."
#| echo: false

nas_prop_bm_draws_long_voicing |> head(n = 10) |> knitr::kable()
```

```{r}
#| label: fig-nas-prop-bm-draws-long-voicing-plot
#| fig-cap: "Expected nasalisation proportion by voicing, averaged across vowels."

@@

The plot suggests an overall greater nasalisation proportion in voiceless NC sequences. Let's quantify how greater as we did in Case Study 1. We will use odd-ratios in this context as well, i.e. we will convert log-odds to odd-ratios using the `exp()` (exponential) function (and as before we first calculate the difference and then exponentiate the resulting values, after which we can take summary measures, like means and quantile-based measures such as CrIs). tbl-nas-prop-bm-diff-quant show the 90, 80 and 60% CrIs of the difference ratio of nasalisation proportion in voiceless vs voiced NC sequences.

@@

nas_prop_bm_diff <- nas_prop_bm_draws_long_voicing |> 
  # pivot data to create one column per voicing with the mean drawn values,
  # with one draw per raw. we need to drop the value_mean_prop col
  select(-value_mean_prop) |> 
  pivot_wider(names_from = voicing, values_from = value_mean) |> 
  mutate(
    # calculate the difference of voiceless and voiced in log-odds
    voicing_diff = voicingvoiceless - voicingvoiced,
    # now transform with exp() to get the ratio difference
    voicing_diff_ratio = exp(voicing_diff)
  )

nas_prop_bm_diff_quant <- nas_prop_bm_diff |> 
  reframe(
    # 90% CrI
    q90 = quantile2(voicing_diff_ratio, probs = c(0.05, 0.95)),
    # 80% CrI
    q80 = quantile2(voicing_diff_ratio, probs = c(0.1, 0.9)),
    # 60% CrI
    q60 = quantile2(voicing_diff_ratio, probs = c(0.2, 0.8)),
  ) |> 
  mutate(across(everything(), ~round(.x, 2)))
```

```{r}
#| label: tbl-nas-prop-bm-diff-quant
#| tbl-cap: "Upper and lower limits of 90, 80 and 60% Credible Intervals of the difference ratio of nasalisation proportion in vowels followed by voiced vs voiceless consonants."
#| echo: false

nas_prop_bm_diff_quant |> knitr::kable()

The CrIs of the ratio difference in nasalisation proportion in voiceless vs voiced NC sequences suggest an increase of nasalisation in the voiceless NC sequences, with a 90% probability that the increase is between 23% and 69% of the proportion in voiced NC sequences.

@@

nas_prop_r <- nas_prop_bm |> 
  # extract draws from model, only `r_speaker` varying terms
  spread_draws(r_speaker[speaker,voicingvowel]) |> 
  # separate the column voicingvowel to two columns
  separate(voicingvowel, c("voicing", "vowel")) |> 
  # join the draws with the `b_` terms
  left_join(y = nas_prop_bm_draws_long) |> 
  # get the expected log-odd value of each speaker, in each draw
  # this is the sum of the `value` from the b_ terms and the value from the
  # r_speaker term.
  mutate(r_speaker_value = value + r_speaker) |> 
  # group the data for summarise
  group_by(.draw, speaker, voicing) |> 
  # get mean expected log-odds by draw, speaker and voicing (averaging across vowel)
  summarise(r_speaker_value_mean = mean(r_speaker_value)) |> 
  # make the data wider: two columns, one for voiced and one for voiceless
  pivot_wider(names_from = voicing, values_from = r_speaker_value_mean) |> 
  # finally, calculate the difference in expected log-odds of voiceless and voiced
  mutate(voicing_diff = voicingvoiceless - voicingvoiced)
```

@@

```{r}
#| label: fig-nas-prop-r
#| fig-cap: "Expected log-odd difference of coarticulatory nasalisation in voiceless vs voiced NC sequences for each speaker."
#| warning: false
#| fig-asp: 1.5
